Network based multifactorial modelling microRNA-target interaction.

### 2019-05-27 article.Rmd

![Shematic presentation of mechanism of network based model. a) In steady state, miRNAs (M: triangle grey) repress targets (G: circle coconut) according to proportion of their targets' expression. b-c-d) MiRNA expression distributions to targets vary when expression of one of the targets perturbed.](Fig1.jpg){#fig1}

The model was developed by making some assumptions. Transcription rates and degradation rates of miRNAs were assumed as stationary and equal. Targets also have stabil transcription and degradation rates, at the same time they are equal. The factors that could be affect the miRNA:target interaction are disregarded in this approach. It is considered that miRNAs degradates or inhibits them via any mechanisms after binding to target while the miRNAs were used reversibly. In initial conditions, gene expression amounts in system have been considered after repression activity of miRNAs. If a change occurs in the system, variation on miRNA proportional distribution are reflected as change of repression activity.
\begin{equation} 
    Eff_{gi}= C_m * C_{gi}/\sum_{1}^{i} C_g \tag{1}\label{eq:1}
\end{equation}
\begin{equation}
   R_{gi}= Eff_{i1}+ Eff_{i2} ... \tag{2}\label{eq:2}
\end{equation}

@@

To determine suitability of method, we have tried to apply model in a real dataset. Firstly, we obtained the miRNA and gene expression datasets of the patient's tumor and normal tissue. We downloaded the high-throughtput experimental datasets which are provided miRNA:gene target pairs with interaction factors by **moore_mirnatarget_2015** and [@helwak_mapping_2013]. We have combined miRNA and gene expression datasets via miRNA:target gene dataset **(Whole steps can be found in Supplementary data(TCGA_E9-A1N5_article.Rmd)**). We selected a gene which is over expressed in tumor tissue with compared to normal tissue. It is pointed out that ABCC1 gene, *Multidrug resistance-associated protein 1*, is one of the most significant factor to develop resistance against chemotherapotic agents[@atalay2008role; @lu2015microrna; @atalay2006multidrug]. So, we selected ABCC1 gene for simulation of integrated dataset. We have detected the iteration that is suitable for ABCC1 gene simulation. After simulation of system via ABCC1 gene with over expression, we have compared simulation results and tumor tissue expression levels.

@@

We have developed a model approach and workflow for competitive ceRNA regulation in this study. The basic mode of miRNA repression activity has been based on miRNA and target abundance in various researches[@arvey_target_2010; @denzler_assessing_2014]. For this reason, we performed first application as considering only abundance of miRNAs and targets (\autoref{fig1}). After the increasing in a gene expression value, expression values of other genes also changed slightly in comparison with alteration in perturbation started gene node. Different studies have shown that the other gene expression values also rise differently, if a gene abundance increases in ceRNA system [@lai_understanding_2016; @salmena_cerna_2011; @tay_multilayered_2014]. It was observed that primary neighborhoods of the trigger gene change with the same ratio, but cooperative efficiency of miRNAs on common target causes differently change in common target (\autoref{fig1}d). The common target in system has displayed like a trigger for the other group and induced changes of expression values of genes on the other group. Therefore, effect of primary change of the gene (G2) on genes of the other group was observed slightly (\autoref{fig1}d). In addition, like shown by ceRNA hypothesis model of Ala et al., after the increasing of a gene expression (G2), the miRNA that is found in the same group (M1) tended to be more repressive on this target. This has caused to decrease in increased expression value of primer triggering gene. When the expression of a gene (G2) was increased to two fold, we observed that the expression changes were more evident. So, we have considered that high miRNA:target ratio results in stronger miRNA repression activity. This results also are coherent with previous reports that were offered by Arvey et al. and Denzler et al.[@arvey_target_2010; @bosson_endogenous_2014; @denzler_assessing_2014]. When the simulation was run in sample dataset that includes lower target abundance, the first response of primary neihgborhood of the same group was determined as similar with in system that have high target abudance. But regulations of ceRNAs were observed differences in the folowing steps of simulation. In overall state in the end of simulation, we have established that all regulations more prominent when compared with low miRNA:target ratio dataset. The model has appeared the importance of mRNA:target ratio in regulation of competing RNAs.

@@

It has been reported in several studies that miRNA regulatory interactions also is affected different parameters. For example, Xu et. al have investigated the importance of seed pairing type between miRNAs and their targets and target site location by using proteomic dataset [@xu_characterization_2014]. They have proposed that the features of binding between miRNA and target can be critical efficiency of miRNA and can be used to find effective miRNA molecules. In addition, binding energy between miRNA and targets also could be determinant for miRNA efficiency. It has been reported by Breda et. al. that strength of miRNA:target interactions is depended on binding energy of complexes [@breda_quantifying_2015]. As similar to these, Bosson et al. have associated affinity and seed pairing of miRNA target pairs and suggested that number of canonical base pairing is correlated with affinity. For these reasons, we have improved the model with optional interaction parameters that could be useful for understanding of miRNA repression activity. We have explained the model in the same sample dataset with using different factors and simulated via the same change (\autoref{fig3}). When the factors were taken into account in the system, miRNA efficiencies varied as shown in \autoref{fig3}a. Although the miRNA target ratios in initial states were same in comparison with the sample dataset without factors, efficiency of binding and repression have changed. When expression of a gene (G2) increased, expression values of all genes also change differentially because of contribution of efficiency factors (\autoref{fig3} b,c,d). We have considered that in our approach energy and seed type of pairs is significant for binding and targeted region is important for repression. But these factors can be handled optionally and differently.

@@

The defined miRNA’ number is very low in comparison with defined protein coding genes’ number. So, this results in that the miRNA amounts in the system must be high or targeted RNAs’ amount must be low for effective miRNA activity. We have tried to simulate a real dataset which contains thousands of genes and hundreds of miRNAs. Firstly, we have observed the activities in the iterations and determined behaviours of the network after simulation. As expected, we have observed that the big dataset, compared to those containing a few genes and miRNAs, exhibited cumbersome behaviours and complexity to gain steady-state **(see details in Supplementary File(TCGA_E9-A1N5_article.Rmd)**. We realized that a small change in a system with a large change actually brought many arrangements. This can be explained with the competing groups. Because the genes are targeted with many common miRNAs, and this results in that the common targets exhibit competing behaviours in different groups. On the other hand, the changes of gene expressions were very different. While the change in some of genes were highly evident, somes have changed slightly. So, we have compared two datasets; the tumor gene expression and simulation results. When we analysed comparison of two datasets, we have observed that the results of simulation is not convenient with tumor tissue expression values. The simulation results were not expected to be concordant with tumor tissue expression values because of a single gene could not be responsible for gene regulation in all tissue. Besides, the all other factors such as up/down regulation of miRNAs or other genes were ignored. Of course, it would be more useful to test a dataset that includes gene expression values before and after the regulation of a gene and miRNA expression values at initial conditions. When we evaluated the suitability of approach, the approach allows different miRNA targets to be considered together and can work with different regulatory elements due to the network structure.

The network based approaches have been developed in previous studies. Figliuzzi et al. have tried to explain the ceRNA crosstalk in a network-like minimal interaction structure with concentrations of ceRNA and miRNAs. They have pointed out that the larger miRNA number can cause to evident crosstalk between ceRNAs. It has been thought that miRNAs and their targets interactions depend on rates of transcription, degradation, binding and unbinding in the network based kinetic model which has been developed by Nitzan et al. Nitzan and collaborators have used the high throughput experimental dataset [@helwak_mapping_2013] for miRNA:target pair that occur high free energy and microarray dataset of a miR-92a depletion for expression values. They have been demonstrated that distant ceRNAs can interact with each other via indirect links, and the interactions change to depend on distance between ceRNAs, and topological features of network [@nitzan_interactions_2014]. List et al. have developed an approach to detect ceRNA interaction by using the miRNA expression, gene expression and common miRNAs between gene targets [@markus_list_sponge_2017]. Their approach could be useful to analysis interacted genes through miRNAs. When it has been throught that a miRNA can exhibit strong functionality to a target but may not against an other, common miRNA based approach can be inconvenient to understand regulations of ceRNA interactions.

>>>>

![Schematic presentation of mechanism of network based model. a) In steady state, miRNAs (M: gray triangle) repress targets (G: circle coconut) according to proportion of their targets' expression. b-c-d) MiRNA expression distributions to targets vary when expression of one of the targets perturbed.](Fig1.jpg){#fig1}

The model was developed by making some assumptions. Transcription rates and degradation rates of miRNAs were assumed as stationary and equal. Targets also have stable transcription and degradation rates, at the same time they are equal. The factors that could be affect the miRNA:target interaction are disregarded in this approach. It is considered that miRNAs degrades or inhibits them via any mechanisms after binding to target while the miRNAs were used reversibly. In initial conditions, gene expression amounts in system have been considered after repression activity of miRNAs. If a change occurs in the system, variation on miRNA proportional distribution are reflected as change of repression activity.
\begin{equation} 
    Eff_{gi}= C_m * C_{gi}/\sum_{1}^{i} C_g \tag{1}\label{eq:1}
\end{equation}
\begin{equation}
   R_{gi}= Eff_{i1}+ Eff_{i2} ... \tag{2}\label{eq:2}
\end{equation}

@@

To determine suitability of method, we have tried to apply model in a real dataset. Firstly, we obtained the miRNA and gene expression datasets of the patient's tumor and normal tissue. We downloaded the high-throughput experimental datasets which are provided miRNA:gene target pairs with interaction factors by **moore_mirnatarget_2015** and [@helwak_mapping_2013]. We have combined miRNA and gene expression datasets via miRNA:target gene dataset **(Whole steps can be found in Supplementary data(TCGA_E9-A1N5_article.Rmd)**). We selected a gene which is over expressed in tumor tissue with compared to normal tissue. It is pointed out that ABCC1 gene, *Multidrug resistance-associated protein 1*, is one of the most significant factor to develop resistance against chemotherapotic agents[@atalay2008role; @lu2015microrna; @atalay2006multidrug]. So, we selected ABCC1 gene for simulation of integrated dataset. We have detected the iteration that is suitable for ABCC1 gene simulation. After simulation of system via ABCC1 gene with over expression, we have compared simulation results and tumor tissue expression levels.

@@

We have developed a model approach and workflow for competitive ceRNA regulation in this study. The basic mode of miRNA repression activity has been based on miRNA and target abundance in various researches[@arvey_target_2010; @denzler_assessing_2014]. For this reason, we performed first application as considering only abundance of miRNAs and targets (\autoref{fig1}). After the increasing in a gene expression value, expression values of other genes also changed slightly in comparison with alteration in perturbation started gene node. Different studies have shown that the other gene expression values also rise differently, if a gene abundance increases in ceRNA system [@lai_understanding_2016; @salmena_cerna_2011; @tay_multilayered_2014]. It was observed that primary neighborhoods of the trigger gene change with the same ratio, but cooperative efficiency of miRNAs on common target causes differently change in common target (\autoref{fig1}d). The common target in system has displayed like a trigger for the other group and induced changes of expression values of genes on the other group. Therefore, effect of primary change of the gene (G2) on genes of the other group was observed slightly (\autoref{fig1}d). In addition, like shown by ceRNA hypothesis model of Ala et al., after the increasing of a gene expression (G2), the miRNA that is found in the same group (M1) tended to be more repressive on this target. This has caused to decrease in increased expression value of primer triggering gene. When the expression of a gene (G2) was increased to two fold, we observed that the expression changes were more evident. So, we have considered that high miRNA:target ratio results in stronger miRNA repression activity. This results also are coherent with previous reports that were offered by Arvey et al. and Denzler et al.[@arvey_target_2010; @bosson_endogenous_2014; @denzler_assessing_2014]. When the simulation was run in sample dataset that includes lower target abundance, the first response of primary neighborhood of the same group was determined as similar with in system that have high target abundance. But regulations of ceRNAs were observed differences in the following steps of simulation. In overall state in the end of simulation, we have established that all regulations more prominent when compared with low miRNA:target ratio dataset. The model has appeared the importance of mRNA:target ratio in regulation of competing RNAs.

@@

It has been reported in several studies that miRNA regulatory interactions also is affected different parameters. For example, Xu et. al have investigated the importance of seed pairing type between miRNAs and their targets and target site location by using proteomics dataset [@xu_characterization_2014]. They have proposed that the features of binding between miRNA and target can be critical efficiency of miRNA and can be used to find effective miRNA molecules. In addition, binding energy between miRNA and targets also could be determinant for miRNA efficiency. It has been reported by Breda et. al. that strength of miRNA:target interactions is depended on binding energy of complexes [@breda_quantifying_2015]. As similar to these, Bosson et al. have associated affinity and seed pairing of miRNA target pairs and suggested that number of canonical base pairing is correlated with affinity. For these reasons, we have improved the model with optional interaction parameters that could be useful for understanding of miRNA repression activity. We have explained the model in the same sample dataset with using different factors and simulated via the same change (\autoref{fig3}). When the factors were taken into account in the system, miRNA efficiencies varied as shown in \autoref{fig3}a. Although the miRNA target ratios in initial states were same in comparison with the sample dataset without factors, efficiency of binding and repression have changed. When expression of a gene (G2) increased, expression values of all genes also change differentially because of contribution of efficiency factors (\autoref{fig3} b,c,d). We have considered that in our approach energy and seed type of pairs is significant for binding and targeted region is important for repression. But these factors can be handled optionally and differently.

@@

The defined miRNA’ number is very low in comparison with defined protein coding genes’ number. So, this results in that the miRNA amounts in the system must be high or targeted RNAs’ amount must be low for effective miRNA activity. We have tried to simulate a real dataset which contains thousands of genes and hundreds of miRNAs. Firstly, we have observed the activities in the iterations and determined behaviors of the network after simulation. As expected, we have observed that the big dataset, compared to those containing a few genes and miRNAs, exhibited cumbersome behaviors and complexity to gain steady-state **(see details in Supplementary File(TCGA_E9-A1N5_article.Rmd)**. We realized that a small change in a system with a large change actually brought many arrangements. This can be explained with the competing groups. Because the genes are targeted with many common miRNAs, and this results in that the common targets exhibit competing behaviors in different groups. On the other hand, the changes of gene expressions were very different. While the change in some of genes were highly evident, some have changed slightly. So, we have compared two datasets; the tumor gene expression and simulation results. When we analyzed comparison of two datasets, we have observed that the results of simulation is not convenient with tumor tissue expression values. The simulation results were not expected to be concordant with tumor tissue expression values because of a single gene could not be responsible for gene regulation in all tissue. Besides, the all other factors such as up/down regulation of miRNAs or other genes were ignored. Of course, it would be more useful to test a dataset that includes gene expression values before and after the regulation of a gene and miRNA expression values at initial conditions. When we evaluated the suitability of approach, the approach allows different miRNA targets to be considered together and can work with different regulatory elements due to the network structure.

The network based approaches have been developed in previous studies. Figliuzzi et al. have tried to explain the ceRNA crosstalk in a network-like minimal interaction structure with concentrations of ceRNA and miRNAs. They have pointed out that the larger miRNA number can cause to evident crosstalk between ceRNAs. It has been thought that miRNAs and their targets interactions depend on rates of transcription, degradation, binding and unbinding in the network based kinetic model which has been developed by Nitzan et al. Nitzan and collaborators have used the high throughput experimental dataset [@helwak_mapping_2013] for miRNA:target pair that occur high free energy and microarray dataset of a miR-92a depletion for expression values. They have been demonstrated that distant ceRNAs can interact with each other via indirect links, and the interactions change to depend on distance between ceRNAs, and topological features of network [@nitzan_interactions_2014]. List et al. have developed an approach to detect ceRNA interaction by using the miRNA expression, gene expression and common miRNAs between gene targets [@markus_list_sponge_2017]. Their approach could be useful to analysis interacted genes through miRNAs. When it has been thought that a miRNA can exhibit strong functionality to a target but may not against an other, common miRNA based approach can be inconvenient to understand regulations of ceRNA interactions.
